### 2017-01-29 Radio/VLA_OH/VLA_2.tex
replace hierarchical with cold gas accretion

\begin{center}
{\large{\bf{Probing OH Megamaser in a Strongly-Lensed Gas-Rich Merger at $z$\ssim0.65}}}
\end{center}
\vspace{-.8em}
\textbf{OH megamasers and the physical conditions of low- and high-$z$ \galpop}
% Galaxy evolution and the history of galaxy mergers}
Studying the properties of high-redshift ($z$) galaxies is crucial to understand the formation and evolution of present-day galaxies.
High-$z$ galaxies are IR-luminous due to their intense bursts
of obscured star formation and/or their central active galactic nuclei (AGNs),
which are sustained by their massive molecular gas reservoirs ($M_{\rm gas}$$>$10$^{10}$\,\Msun; see \citealt{CW13} for a review).
The gas-rich and starbursting nature of high-$z$ galaxies may be a result of galaxy mergers
\citep{Riechers11a, Hayward12a, Riechers13b},
consistent with the notion that galaxies grow through hierarchical merging \citep{White91a}.
%  merging in the hierarchical model of galaxy formation \citep{White91a}.
The merger rate is thus expected to increase towards higher redshift \citep[e.g.,][]{LeFloch05a, Magnelli09a}, and constraining
its evolution allows us to track the origin and evolution of present-day populations. % emergence of current day populations.

\noindent The steep dependence between OH and FIR luminosities ($L_{\rm OH}^2\propto L_{\rm FIR}^2$; \Fig{model}a)
found in local OH ``megamasers'' (OHMs;  see \citealt[]{Lo05a} for a review) and
the ubiquity of strong IR-emitting galaxies found at high $z$ suggest that
the number density of OHM may improve current constraints on
the cosmic merger history \citep[hereafter DG02]{Darling02b_LF}.
Existing models predict that detections of
OHMs out to at least $z$\ssim3 are possible with current facilities (e.g. the GBT; \Fig{model}b).
Yet, despite several attempts,
the most distant OHM known-to-date was discovered over two decades ago, at $z$\eq0.265 \citep{Baan92a}.
The null detection in a $z$\ssim4 extremely IR-bright galaxy (\LIR$>10^{14}$\Lsun; \citealt{Ivison06a}) and
the remarkably low detection rate (7\%) from a sample of $>$120 IR-luminous galaxies
at $z$\,$\lesssim$\,1.5 ($L_{\rm IR}>10^{11}$\Lsun; \Fig{model}b; \citealt{Willett12a})
suggest a potential redshift evolution in the \LOH-\LFIR relation.
Detecting OHM beyond $z$\eq0.265 is therefore needed to enable a better understanding of the
OHM trigger mechanism, which may differ at earlier epochs due to different physical conditions of galaxies,
and to improve existing predictions to plan for future OHM surveys with upcoming facilities
\citep[e.g., FAST, APERTIF/WSRT, ASKAP;][]{Zhang14b}.
Here we propose to observe OH 1667 MHz line emission in the
gravitationally-lensed starbursting galaxy merger
RXJ1131 at $z$\eq0.65 (\Fig{hst}) using the VLA's B-array configuration.

\noindent \textbf{An ideal target --- a lensed gas-rich, starbursting merger at $z$\,$>$\,0.6}\\
\noindent High-resolution imaging and spectroscopy in optical wavelength show that RXJ1131 is
lensed by a foreground galaxy at $z$\ssim0.3 (\Fig{hst}; \citealt{Sluse03a}).
By exploiting the effect of lensing magnification,
our proposed detection experiment requires {\bf an on-source time of  % sigma scales as 1/root(t) --> t scales as 1/sigma^2
$\sim$30 times less} than otherwise needed to reach the same sensitivity.
The unique lensing configuration of RXJ1131 has led to extensive follow-up studies spanning X-ray to radio \citep[\Fig{hst}e; e.g.,][]{Claeskens06a, Sluse07a, Leung17a},
making it one of the best-studied galaxies at $z$$>$0.5 and
the only source at 0.2$<$$z$$<$1 with spatially-resolved imaging of the molecular gas distribution \citep{Leung17a}.
% X-ray studies  a highly compact source of emission
% in the innermost region near the accretion disk down to pc-scale
% \citep[$\sim$20\,AU][]{Dai10a}.
Modeling of the X-ray spectrum reveals substantial amount of neutral gas with column density of
 % The continuum is strongly absorbed at soft energies by a
$N_{\rm H}\sim$10$^{23}$ cm$^{-2}$ in the innermost region near % the accretion disk % w.in the X-ray emitting region
% The spectrum at lower energies is heavily absorbed by neutral gas with a hydrogen column density $N_{\rm H}$.
% Soft X-ray continuum suggest the presence of a substantial amount of gas.
the $L_X$$\gtrsim$10$^{45}$\,\Lsun AGN \citep{Pooley07a, Reis14a}.
Our best-fit spectral energy distribution (SED) model indicates
a strong apparent far-IR luminosity of \LFIR$\sim$5\E{12}\Lsun, which corresponds to
an intense star formation rate of SFR\ssim120 \Msun yr\pmOne (lensing-corrected).
The large column density together with the strong FIR radiation field % of $\sim$10$^{12}$\Lsun (lensing-corrected),
% recent star formation (as traced by rest-frame UV emission),
% presence of a nearby companion, and % at $\sim$2.4\,kpc away, and
and the large molecular gas reservoir of $M_{\rm gas}$\,$\gtrsim$\,10$^{10}$\Msun (\Fig{hst}b, c)
are favourable for generating OH masers, making RXJ1131 an ideal candidate
to search for OHMs beyond the nearby universe. Assuming the Malmquist-bias corrected relation
($L_{\rm OH}\propto L_{\rm FIR}^{1.2}$; DG02), we expect an apparent OH luminosity of \LOH= 3.3\E{3}\,\Lsun (orange marker in \Fig{model}b).

\noindent In addition, the redshift of RXJ1131 has been spectroscopically confirmed with CO lines down to a precision of
$\Delta z$\eq0.0002, indicating that the proposed OH line is redshifted to a fortuitous region
free of RFI at the low-end of the L-band.
Together with the high fringe rate of the B-array configuration, the proposed \obs will
reach a sensitivity deeper than previous high-$z$ OHM searches beyond $z$$>$0.4, which were carried out
with single-dish telescopes \citep[\Fig{model}b; e.g.][]{Darling02a, Willett12a}.

\noindent \textbf{Science goals}
\vspace{.2em}\\
$\bullet$ {\bf Testing OHM as a tracer of merger}
The proposed \obs will establish the first OHM measurement in a {\it bona-fide} high-$z$ merger, thereby
demonstrating the feasibility of using it as a tracer of distant mergers and
testing the long-envisaged hypothesis of using its number density to constrain
the redshift evolution of merger rate.
Such a detection will also allow us to better constrain the
\LOH-\LFIR relation out to higher $z$, % which may provide a handle on the pumping efficiency (\LOH/\LFIR)
% that may be closer to other high-$z$ galaxies if their physical conditions are more similar than compared to low-z (\LOH/\LFIR)
and thus improve predictions on the required sensitivity needed for future searches of high-$z$ OHMs.
Since our target is a merger,
a null detection would also indicate that the notion of
using the OHM number density to constrain the merger history requires revision before designing future OHM surveys.

%
% OHM help us understand the physical conditions of high-z galaxies: different circumnuclear environment, optical depth? - likely already know about the cirucmnuclear env part from X-ray, but then could be use for other high-z sources w/o detailed characterization?
\noindent $\bullet$ {\bf Physical conditions and OH pumping mechanisms in low- and high-$z$ galaxies}
% Classical model: foreground amplification of background radio continuum, expect quadratic relationship, if the population inversion is pumped by the FIR radiation field and the radio continuum (which itself is proportional to the FIR luminosity) is being amplified by an unsaturated maser cloud.
Null detections in previous high-$z$ OHM searches suggest a shallower \LOH-\LFIR dependence, which may be
reflecting the different physical conditions of the inner few hundred pc (where OHMs are generated)
between low- and high-$z$ IR-luminous galaxies.
The lower \LOH/\LFIR indicate that the OH amplification is less efficient at the same FIR luminosity
and that collisional pumping (e.g. due to the strong AGN, and high gas mass and density; \citealt{Field94a}) % volume density
% while locally, not all OHM are AGN hosts, suggesting the FIR radiation from the SB is the dominant pumping source, but the non-detection based on local extrapolation and such maser model may suggest a different energy source/powering mechanism at higher-z, where quasars are more active, and may indicate that something about AGN/radio pumping
%
% Since a purely radiative pumping mechanism is possible
% only at relatively low densities ($n_{\rm OH}$\ssim10$^{-4}$\,cm$^{-3}$), and collisional processes can theoretically predominate at high OH densities, and saturated amplification can become important \citep{Field94a}.
may play a dominant role in these gas-rich \highz systems.
Obtaining a definitive OHM measurement in RXJ1131 will provide a handle on the OH pumping efficiency (\LOH/\LFIR) out to higher $z$.
This together with the detailed properties derived for our target (e.g., \LFIR, \LIR, SFR, $L_{\rm radio}$, molecular and dust content)
will elucidate the role of collisional over radiative pumping
that may be prevalent in \highz galaxies given their physical properties, which are more similar to RXJ1131 than
local starbursts. % more violent star formation in a more gas-rich turbulent ISM.
	% [NOT applicable] Since the optical depth increases with FIR radiation field, the simple picture of low-gain amplification
	% (i.e., unsaturated masing) found to be consistent with local OHMs...
	%
% % implying the simple picture of low-gain amplification (i.e., unsaturated masing, quadrature, low density)  \citep{Henkel90a,Randell95a} found to be consistent with local OHMs will likely break down; (i.e., not all OHMs are unsaturated).
% While in the nearby universe (z\ssim0), OHMs appear to be unsaturated.

\noindent $\bullet$ {\bf OHM as a tracer of the kinematics in the innermost region of distant galaxies}
The much broader linewidths found in $z$$>$0.2 OHMs compared to those at $z$\eq0 already indicate
different dynamical properties in the innermost regions of their hosts \citep{Baan92a,Darling02a}.
Detecting the OHM in RXJ1131 will therefore demonstrate the feasibility of
using it as a trace of the kinematics in the circumnuclear regions of \highz galaxies, which will improve
our understanding of the internal dynamics of IR-luminous, starbursting galaxies at high $z$.

\noindent $\bullet$ {\bf Radio continuum}
In addition, the proposed \obs will provide an additional constraint on the radio continuum emission at 1\,GHz
towards RXJ1131 (\Fig{hst}e).

>>>>

\begin{center}
{\large{\bf{Probing OH Megamaser in a Strongly-Lensed Gas-Rich Merger at $z$\ssim0.65}}}
\end{center}
\vspace{-.8em}
\textbf{OH megamasers and the physical conditions of low- and high-$z$ \galpop}
Studying the properties of high-redshift ($z$) galaxies is crucial to understand the formation and evolution of present-day galaxies.
High-$z$ galaxies are IR-luminous due to their intense bursts
of obscured star formation and/or their central active galactic nuclei (AGNs),
which are sustained by their massive molecular gas reservoirs ($M_{\rm gas}$$>$10$^{10}$\,\Msun; see \citealt{CW13} for a review).
% Their high IR luminosity imply that they are responsible for building most of the stellar mass in the universe.
The gas-rich and starbursting nature of high-$z$ galaxies may be a result of galaxy mergers
\citep{Riechers11a, Hayward12a, Riechers13b}, or cold gas accretion from the cosmic web \citep[e.g.,][]{Dekel09b}.
% , consistent with the notion that galaxies grow through hierarchical merging \citep{White91a}.
Since the merger rate is expected to increase towards higher redshift \citep[e.g.,][]{LeFloch05a, Magnelli09a}, constraining
its evolution allows us to track the origin and evolution of present-day populations. % emergence of current day populations.

%Galaxies with high IR-luminous and intense star formation are important for the build-up of stellar mass in the universe. 
%While IR-luminous galaxies are found in both nearby and at high $z$, studies find that they display different properties and 
%are different galaxy populations. Low-$z$ ULIRGs are predominantly more compact,
%mostly merger-triggered starbursts, with lower molecular 
%gas mass fractions \citep{Veilleux02a, Hopkins10a, Rujopakarn11a, Rujopakarn13a}, whereas 
%\highz galaxies are a combination of mergers and cold gas accretion.

\noindent The steep dependence between OH and FIR luminosities ($L_{\rm OH}^2\propto L_{\rm FIR}^2$; \Fig{model}a)
found in local OH ``megamasers'' (OHMs;  see \citealt[]{Lo05a} for a review) and
the ubiquity of strong IR-emitting galaxies found at high $z$ suggest that
the number density of OHM may improve current constraints on
the cosmic merger history \citep[hereafter DG02]{Darling02b_LF}. % but that's assuming IR-luminous must be driven by mergers....
Existing models predict that detections of
OHMs out to at least $z$\ssim3 are possible with current facilities (e.g. the GBT; \Fig{model}b).
Yet, despite several attempts,
the most distant OHM known-to-date was discovered over two decades ago, at $z$\eq0.265 \citep{Baan92a}.
The null detection in a $z$\ssim4 extremely IR-bright galaxy (\LIR$>10^{14}$\Lsun; \citealt{Ivison06a}) and
the remarkably low detection rate (7\%) from a sample of $>$120 IR-luminous galaxies
at $z$\,$\lesssim$\,1.5 ($L_{\rm IR}>10^{11}$\Lsun; \Fig{model}b; \citealt{Willett12a})
suggest a potential redshift evolution in the \LOH-\LFIR relation.
Detecting OHM beyond $z$\eq0.265 is therefore needed to enable a better understanding of the
OHM trigger mechanism, which may differ at earlier epochs due to different physical conditions of galaxies,
and to improve existing predictions to plan for future OHM surveys with upcoming facilities
\citep[e.g., FAST, APERTIF/WSRT, ASKAP;][]{Zhang14b}.
Here we propose to observe OH 1667 MHz line emission in the
gravitationally-lensed starbursting galaxy merger
RXJ1131 at $z$\eq0.65 (\Fig{hst}) using the VLA's B-array configuration.

\noindent \textbf{An ideal target --- a lensed gas-rich, starbursting merger at $z$\,$>$\,0.6}\\
\noindent High-resolution imaging and spectroscopy in optical wavelength show that RXJ1131 is
lensed by a foreground galaxy at $z$\ssim0.3 (\Fig{hst}; \citealt{Sluse03a}).
The unique lensing configuration of RXJ1131 has led to extensive follow-up studies spanning X-ray to radio \citep[\Fig{hst}e; e.g.,][]{Claeskens06a, Sluse07a, Leung17a},
making it one of the best-studied galaxies at $z$$>$0.5 and
the only source at 0.2$<$$z$$<$1 with spatially-resolved imaging of the molecular gas distribution \citep{Leung17a}.
% X-ray studies  a highly compact source of emission
% in the innermost region near the accretion disk down to pc-scale
% \citep[$\sim$20\,AU][]{Dai10a}.
Modeling of the X-ray spectrum reveals substantial amount of neutral gas with column density of
 % The continuum is strongly absorbed at soft energies by a
$N_{\rm H}\sim$10$^{23}$ cm$^{-2}$ in the innermost region near % the accretion disk % w.in the X-ray emitting region
% The spectrum at lower energies is heavily absorbed by neutral gas with a hydrogen column density $N_{\rm H}$.
% Soft X-ray continuum suggest the presence of a substantial amount of gas.
the $L_X$$\gtrsim$10$^{45}$\,\Lsun AGN \citep{Pooley07a, Reis14a}.
Our best-fit spectral energy distribution (SED) model indicates
a strong apparent far-IR luminosity of \LFIR$\sim$5\E{12}\Lsun, which corresponds to
an intense star formation rate of SFR\ssim120 \Msun yr\pmOne (lensing-corrected).
The large column density together with the strong FIR radiation field % of $\sim$10$^{12}$\Lsun (lensing-corrected),
% recent star formation (as traced by rest-frame UV emission),
% presence of a nearby companion, and % at $\sim$2.4\,kpc away, and
and the large molecular gas reservoir of $M_{\rm gas}$\,$\gtrsim$\,10$^{10}$\Msun (\Fig{hst}b, c)
are favourable for generating OH masers, making RXJ1131 an ideal candidate
to search for OHMs beyond the nearby universe.

\noindent In addition, the redshift of RXJ1131 has been spectroscopically confirmed with CO lines down to a precision of
$\Delta z$\eq0.0002, indicating that the proposed OH line is redshifted to a fortuitous region
free of RFI at the low-end of the L-band.
Together with the high fringe rate of the B-array configuration, the proposed \obs will
reach a sensitivity deeper than previous high-$z$ OHM searches beyond $z$$>$0.4, which were carried out
with single-dish telescopes \citep[\Fig{model}b; e.g.][]{Darling02a, Willett12a}.
Assuming the Malmquist-bias corrected relation
($L_{\rm OH}\propto L_{\rm FIR}^{1.2}$; DG02), we expect an apparent OH luminosity of \LOH= 3.3\E{3}\,\Lsun (orange marker in \Fig{model}b). 
We here aim to detect the OHM in RXJ1131 at 5$\sigma$. 
By exploiting the effect of lensing magnification, we require an {\bf on-source time of 4.3 hours}, which
is {\bf $\sim$30 times cheaper}
 % sigma scales as 1/root(t) --> t scales as 1/sigma^2
than otherwise possible.

\noindent \underline{\textbf{Science goals}}
\vspace{.2em}\\
$\bullet$ {\bf Testing OHM as a tracer of merger.}
The proposed \obs will establish the first OHM measurement in a {\it bona-fide} high-$z$ merger, thereby
demonstrating the feasibility of using it as a tracer of distant mergers and
testing the long-envisaged hypothesis of using its number density to constrain
the redshift evolution of merger rate.
Such a detection will also allow us to better constrain the
\LOH-\LFIR relation out to higher $z$, % which may provide a handle on the pumping efficiency (\LOH/\LFIR)
% that may be closer to other high-$z$ galaxies if their physical conditions are more similar than compared to low-z (\LOH/\LFIR)
and thus improve predictions on the required sensitivity needed for future searches of high-$z$ OHMs.
Since our target is a merger,
a null detection would also indicate that the notion of
using the OHM number density to constrain the merger history requires revision before designing future OHM surveys.

%
% OHM help us understand the physical conditions of high-z galaxies: different circumnuclear environment, optical depth? - likely already know about the cirucmnuclear env part from X-ray, but then could be use for other high-z sources w/o detailed characterization?
\noindent $\bullet$ {\bf Physical conditions and OH pumping mechanisms in low- and high-$z$ galaxies.}
% Classical model: foreground amplification of background radio continuum, expect quadratic relationship, if the population inversion is pumped by the FIR radiation field and the radio continuum (which itself is proportional to the FIR luminosity) is being amplified by an unsaturated maser cloud.
Null detections in previous high-$z$ OHM searches suggest a shallower \LOH-\LFIR dependence, which may be
reflecting the different physical conditions of the inner few hundred pc (where OHMs are generated)
between low- and high-$z$ IR-luminous galaxies.
The lower \LOH/\LFIR indicate that the OH amplification is less efficient at the same FIR luminosity
and that collisional pumping (e.g. due to the strong AGN, and high gas mass and density; \citealt{Field94a}) % volume density
% while locally, not all OHM are AGN hosts, suggesting the FIR radiation from the SB is the dominant pumping source, but the non-detection based on local extrapolation and such maser model may suggest a different energy source/powering mechanism at higher-z, where quasars are more active, and may indicate that something about AGN/radio pumping
%
% Since a purely radiative pumping mechanism is possible
% only at relatively low densities ($n_{\rm OH}$\ssim10$^{-4}$\,cm$^{-3}$), and collisional processes can theoretically predominate at high OH densities, and saturated amplification can become important \citep{Field94a}.
may play a dominant role in these gas-rich \highz systems.
Obtaining a definitive OHM measurement in RXJ1131 will provide a handle on the OH pumping efficiency (\LOH/\LFIR) out to higher $z$.
This together with the detailed properties derived for our target (e.g., \LFIR, \LIR, SFR, $L_{\rm radio}$, molecular and dust content)
will elucidate the role of collisional over radiative pumping
that may be prevalent in \highz galaxies given their physical properties, which are more similar to RXJ1131 than
local starbursts. % more violent star formation in a more gas-rich turbulent ISM.
	% [NOT applicable] Since the optical depth increases with FIR radiation field, the simple picture of low-gain amplification
	% (i.e., unsaturated masing) found to be consistent with local OHMs...
	%
% % implying the simple picture of low-gain amplification (i.e., unsaturated masing, quadrature, low density)  \citep{Henkel90a,Randell95a} found to be consistent with local OHMs will likely break down; (i.e., not all OHMs are unsaturated).
% While in the nearby universe (z\ssim0), OHMs appear to be unsaturated.

\noindent $\bullet$ {\bf OHM as a tracer of the kinematics in the innermost region of distant galaxies.}
The much broader linewidths found in $z$$>$0.2 OHMs compared to those at $z$\eq0 already indicate
different dynamical properties in the innermost regions of their hosts \citep{Baan92a,Darling02a}.
Detecting the OHM in RXJ1131 will therefore demonstrate the feasibility of
using it as a trace of the kinematics in the circumnuclear regions of \highz galaxies, which will improve
our understanding of the internal dynamics of IR-luminous, starbursting galaxies at high $z$.

\noindent $\bullet$ {\bf Radio continuum.}
In addition, the proposed \obs will provide an additional constraint on the radio continuum emission at 1\,GHz
towards RXJ1131 (\Fig{hst}e).
